---
title: "Storm and Severe Weather threaden Public Health and Economic Problems"
output: github_document
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This investigation provides answers to the following two questions:

1) Across the United States, which types of events are most harmful with respect to population health?

2) Across the United States, which types of events have the greatest economic consequences?



## Data Download & Ingestion
```{r, echo = TRUE, eval = TRUE, cache=TRUE}
# set work directory
setwd("/Users/joergheintz/Documents/08_MPHPHI/11_Coursera/Coursera_ReproduciableResearch/RepResProj2")

# download the data set from the source
#mypath<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"


#download.file(mypath,"myfile.bz2")
mydata<-read.csv("myfile.bz2")[,c("STATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", "BGN_DATE")]


# Split date from time, keep date
BGN_DATE<-data.frame(t(as.data.frame(strsplit(as.character(mydata$BGN_DATE),' ')))[,1], row.names=NULL)
mydata$BGN_DATE<-as.Date(mydata$BGN_DATE, "%m/%d/%Y")
mydata$FATALITIES<-as.numeric(as.character(mydata$FATALITIES))


## split data set into Fatalities/Injuiries and EconomicDamage 
# Fatalities & Injuries
df_FAT_INJ<-mydata[mydata$FATALITIES != 0 | mydata$INJURIES != 0 ,c("STATE", "EVTYPE", "FATALITIES", "INJURIES", "BGN_DATE")]
write.csv(df_FAT_INJ, "FAT_INJ.csv")


# Economic Damage
df_ECO_DAM<-mydata[mydata$PROPDMG != 0 | mydata$PROPDMG != 0 | mydata$CROPDMG != 0 | mydata$CROPDMGEXP!=0, c("STATE", "EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", "BGN_DATE") ] 
write.csv(df_ECO_DAM, "ECO_DAM.csv")

```


## Fatilities
The data set is clean from all columns that are not needed, and documented in table.  
```{r, echo = TRUE, eval = TRUE}
# Ranking by event Top 10
myFAT<-read.csv("FAT_INJ.csv")
FAT<-as.data.frame(tapply(myFAT$FATALITIES, myFAT$EVTYPE, sum))
FAT$EVENT<-rownames(FAT)
colnames(FAT)<-c("fatalities", "events")
rownames(FAT)<-NULL
FAT<-FAT[order(-FAT$fatalities),]
FAT10<-FAT[1:10,]
FAT10


# sum up over time
fat<-data.frame()
for (ev in FAT10$events){
        sf<-myFAT[myFAT$EVTYPE == ev, ]
        sf<-sf[order(as.Date(sf$BGN_DATE, format="%Y-%m-%d")),]
        sf$SumFATALITIES<-sf$FATALITIES
        k<-length(sf$FATALITIES)-1
        for (i in 1:k){
                sf$SumFATALITIES[i+1] = sf$SumFATALITIES[i]+sf$SumFATALITIES[i+1]
                }
        k=0
        fat<-rbind(fat,sf)
}


#fat<-fat[fat$FATALITIES!=0,]
fat$BGN_DATE<-fat$BGN_DATE<-as.Date(fat$BGN_DATE, format="%Y-%m-%d")
fat<-fat[order(as.Date(fat$BGN_DATE, format="%Y-%m-%d")),]


library(lattice)
library(ggplot2)
g = ggplot(fat, aes(x =BGN_DATE , y = SumFATALITIES, group = EVTYPE, color = factor(EVTYPE)))
g = g + xlab("Date") + ylab("Fatality Development")
g = g + geom_line()
#g = g + geom_point( size=1, shape=21, fill="white")
g = g + scale_x_date(date_breaks ="10 year")
g


```

The turnado is regarding fatility the most deadly event. 

## Injuries
The data set is clean from all columns that are not needed, and documented in table.  
```{r, echo = TRUE, eval = TRUE}
# Ranking by event Top 10
myInj<-read.csv("FAT_INJ.csv")
INJ<-as.data.frame(tapply(myFAT$INJURIES, myFAT$EVTYPE, sum))
INJ$EVENT<-rownames(INJ)
colnames(INJ)<-c("injuries", "events")
rownames(INJ)<-NULL
INJ<-INJ[order(-INJ$fatalities),]
INJ10<-INJ[1:10,]
INJ10

# sum up over time
fat<-data.frame()
for (ev in FAT10$events){
        sf<-myFAT[myFAT$EVTYPE == ev, ]
        sf<-sf[order(as.Date(sf$BGN_DATE, format="%Y-%m-%d")),]
        sf$SumFATALITIES<-sf$FATALITIES
        k<-length(sf$FATALITIES)-1
        for (i in 1:k){
                sf$SumFATALITIES[i+1] = sf$SumFATALITIES[i]+sf$SumFATALITIES[i+1]
                }
        k=0
        fat<-rbind(fat,sf)
}


#fat<-fat[fat$FATALITIES!=0,]
fat$BGN_DATE<-fat$BGN_DATE<-as.Date(fat$BGN_DATE, format="%Y-%m-%d")
fat<-fat[order(as.Date(fat$BGN_DATE, format="%Y-%m-%d")),]


library(lattice)
library(ggplot2)
g = ggplot(fat, aes(x =BGN_DATE , y = SumFATALITIES, group = EVTYPE, color = factor(EVTYPE)))
g = g + xlab("Date") + ylab("Fatality Development")
g = g + geom_line()
#g = g + geom_point( size=1, shape=21, fill="white")
g = g + scale_x_date(date_breaks ="10 year")
g

